Open archaeology, open source?

### 2024-02-01 analysis/_06-network.qmd
Worked on the networks section.
Added more tangible insights to the text, experimented with new layouts (repo-user looks good now), refactored the way the repo-user graph is generated.

# Repo-user
oarch_graph_repo_user <- tbl_graph(nodes_repos_users(oarch_graph),
                                   edges_user_to_repo(oarch_graph))

@@

# User-user
tbl_graph(nodes = nodes_users(oarch_graph), 
          edges = edges_common_repos(oarch_graph),
          directed = FALSE) |>
  mutate(
    degree = centrality_degree(),
    betweenness = centrality_betweenness(weights = n),
    subgraph = as_factor(group_components())
  ) ->
  oarch_graph_user_user

@@

```{r fig-graph-repo-user}
#| fig-cap: Interactive repository–user graph.
# TODO: is there a khroma scale that can accomodate this many groups?
g_repo_user <- oarch_graph_repo_user |>
  ggraph(layout = "graphopt", mass = 100) +
  geom_edge_link(aes(group = contrib, alpha = n_contrib)) +
  geom_point_interactive(aes(x, y, colour = node_type, tooltip = node)) + 
  scale_edge_alpha(trans = "log", guide = guide_none()) +
  scale_colour_discrete(guide = guide_none()) +
  labs(edge_alpha = "Common contributors") +
  theme(panel.background = element_rect(fill = "white"))

@@

```{r fig-graph-repo-repo}
#| fig-cap: Interactive repository–repository graph. Repositories are connected by contributions by common users, with darker edges indicating more common contributors. Node colour indicates cluster membership according to the edge–betweenness method [@girvan2002]. Excludes isolate nodes.
# TODO: is there a khroma scale that can accomodate this many groups?
g_repo_repo <- oarch_graph_repo_repo |>
  ggraph(layout = "graphopt", mass = 100) +
  geom_edge_link(aes(alpha = n)) +
  geom_point_interactive(aes(x, y, colour = group, tooltip = repo)) + 
  scale_edge_alpha(trans = "log", guide = guide_none()) +
  scale_colour_discrete(guide = guide_none()) +
  labs(edge_alpha = "Common contributors") +
  theme(panel.background = element_rect(fill = "white"))

@@

```{r fig-graph-user-user}
#| fig-cap: Interactive user–user graph. Users are connected by contributions by common repositories, with darker edges indicating more common repositories. Node colour indicates cluster membership according to the edge–betweenness method [girvan2002]. Excludes isolate nodes.
# TODO: is there a khroma scale that can accomodate this many groups?
g_user_user <- oarch_graph_user_user |>
  filter(degree > 170) |>
  ggraph(layout = "graphopt", mass = 100) +
  geom_edge_link(aes(alpha = n)) +
  geom_point_interactive(aes(x, y, tooltip = user)) + 
  scale_edge_alpha(trans = "log", guide = guide_none()) +
  scale_colour_discrete(guide = guide_none()) +
  labs(edge_alpha = "Common repositories") +
  theme(panel.background = element_rect(fill = "white"))

@@

<!-- Is there a network? How big is it? -->
Our data shows that there is a significant network of archaeologists collaborating on GitHub.
`r percent(graph_n_repos / graph_n_repos_total)` of repositories and `r percent(graph_n_users / graph_n_users_total)` of users in our dataset are connected to at least one other repository or user.
Of these, `r percent(max(subgraphs_n_repos$n) / graph_n_repos)` of repositories and `r percent(max(subgraphs_n_users$n) / graph_n_users)` of users belong to a single connected subgraph (@fig-graph-repo-repo).

<!-- Clustering within the network -->
To explore the topography of this network, we applied the edge–betweenness community detection method [@girvan2002] to delimit `r nrow(filter(clusters_n_repos, n > 1))` distinct clusters in the repository–repository network.
While many of these clusters are interconnected, some discrete components containing between 2-20 repositories appear as distinct from a primary core.
The core cluster is characterized by repositories whose contributors commit to projects other than their own, and it includes a smorgasbord of projects whose contributors share varied interests.
This indicates the formation of a central software development community rather than a focus on any specific topic of work.
The other peripheral clusters situated around the core largely correspond with the work of single individuals, and sometimes also their close colleagues.
Many of the projects represented in the core emerge from underlying professional partnerships, namely research labs (e.g. ISAA-Kiel) and special interest groups (SSLA).
Peripheral clusters that are connected to core clusters by only a few relationships represent the sole (or perhaps initial) integration of lone developers into a broader community.

@@

The highest mean centrality is seen in repositories that contain lists/datasets, standalone software, or packages/libraries, and in repositories based on Python, R or QGIS [@fig-repo-betweenness-category-platform].
Interestingly, these trends depart from the observed popularity of different categories/platform in the open-archaeo dataset as a whole (see @sec-open-archaeology): standalone software is more central than packages/libraries, even though there are more of the latter by a significant margin.
Similarly, Python packages tend to be more central than R packages, despite being the minority language.
<!-- @ZB: I think this is interesting, but I couldn't really think of a general observation to make here. Maybe you can? -->
<!-- Possibly that R packages are easier to spin up solo, whereas standalone tends to have more financial/institutional backing, and thus greater potential to support teams of developers -->

>>>>

# Repo-user
tbl_graph(nodes_repos_users(oarch_graph),
          edges_user_to_repo(oarch_graph),
          directed = FALSE) |>
          mutate(
            degree = centrality_degree(),
            betweenness = centrality_betweenness(),
            subgraph = as_factor(group_components()),
            group = as_factor(group_edge_betweenness())
            ) ->
            oarch_graph_repo_user

@@

# User-user
tbl_graph(nodes = nodes_users(oarch_graph), 
          edges = edges_common_repos(oarch_graph),
          directed = FALSE) |>
  mutate(
    degree = centrality_degree(),
    betweenness = centrality_betweenness(weights = n),
    subgraph = as_factor(group_components())
    # group = as_factor(group_edge_betweenness()) # <- very slow!
  ) ->
  oarch_graph_user_user

@@

```{r fig-graph-repo-user, eval=FALSE}
#| fig-cap: Interactive repository–user graph.
# TODO: is there a khroma scale that can accomodate this many groups?
g_repo_user <- oarch_graph_repo_user |>
  ggraph(layout = "fr", repulserad = 10, area = 10000, niter = 5000, grid = "nogrid") +
  geom_edge_link(aes(group = contrib, alpha = n_contrib)) +
  geom_point_interactive(aes(x, y, colour = group, shape=node_type, tooltip = node)) + 
  scale_edge_alpha(trans = "log", guide = guide_none()) +
  scale_colour_discrete(guide = guide_none()) +
  labs(edge_alpha = "Common contributors") +
  theme(panel.background = element_rect(fill = "white"))

@@

```{r fig-graph-repo-repo, eval=FALSE}
#| fig-cap: Interactive repository–repository graph. Repositories are connected by contributions by common users, with darker edges indicating more common contributors. Node colour indicates cluster membership according to the edge–betweenness method [@girvan2002]. Excludes isolate nodes.
# TODO: is there a khroma scale that can accomodate this many groups?
g_repo_repo <- oarch_graph_repo_repo |>
  ggraph(layout = "graphopt", mass = 100) +
  geom_edge_link(aes(alpha = n)) +
  geom_point_interactive(aes(x, y, colour = group, tooltip = repo)) + 
  scale_edge_alpha(trans = "log", guide = guide_none()) +
  scale_colour_discrete(guide = guide_none()) +
  labs(edge_alpha = "Common contributors") +
  theme(panel.background = element_rect(fill = "white"))

@@

```{r fig-graph-user-user, eval=FALSE}
#| fig-cap: Interactive user–user graph. Users are connected by contributions by common repositories, with darker edges indicating more common repositories. Node colour indicates cluster membership according to the edge–betweenness method [girvan2002]. Excludes isolate nodes.
# TODO: is there a khroma scale that can accomodate this many groups?
g_user_user <- oarch_graph_user_user |>
  filter(degree > 170) |>
  ggraph(layout = "fr", repulserad = 10, area = 100000, niter = 10000000, grid = "nogrid") +
  geom_edge_link(aes(alpha = n)) +
  geom_point_interactive(aes(x, y, tooltip = user)) + 
  scale_edge_alpha(trans = "log", guide = guide_none()) +
  scale_colour_discrete(guide = guide_none()) +
  labs(edge_alpha = "Common repositories") +
  theme(panel.background = element_rect(fill = "white"))

@@

<!-- Is there a network? How big is it? -->
Our data shows that there is a significant network of archaeologists collaborating on GitHub.
`r percent(graph_n_repos / graph_n_repos_total)` of repositories and `r percent(graph_n_users / graph_n_users_total)` of users in our dataset are connected to at least one other repository or user.
Of these, `r percent(max(subgraphs_n_repos$n) / graph_n_repos)` of repositories and `r percent(max(subgraphs_n_users$n) / graph_n_users)` of users belong to a single connected subgraph (@fig-graph-repo-repo and @fig-graph-user-user).

<!-- Clustering within the network -->
To explore the topography of this network, we applied the edge–betweenness community detection method [@girvan2002] to delimit `r nrow(filter(clusters_n_repos, n > 1))` distinct clusters in the repository–repository network.
While many of these clusters are interconnected, some discrete components containing between 2-20 repositories appear as distinct from a primary core.
The core cluster is characterized by repositories whose contributors commit to projects other than their own, and it includes a smorgasbord of projects whose contributors share varied interests.

Likewise, we applied the same edge-betweenness community detection method to identify clusters in the user-user network, which reveal collaborative groupings.
We again see a complementary primary core connected to several more peripheral clusters, which are internally-cohesive and exhibit few connections with other peripheral clusters.
The central core bridges all the peripheral clusters.
The central core is not uniform, and comprises several relatively discrete sub-groups representing collaborative sub-communities.
While these sub-groups are internally cohesive, they exhibit enough connections to other members of the central core so as to not be considered as separate or peripheral clusters.

In both the repository-repository and user-user networks, the peripheral clusters correspond with either the connections surrounding specific projects or the series of repositories created by single individuals and sometimes also their close colleagues.
On the other hand, the central cores exhibit greater internal variety that may correspond with social connections and the formation of a complex software development community.
This is evident through the fact that many of the connections represented in the cores emerge from underlying professional partnerships, namely research labs (e.g. ISAA-Kiel) and special interest groups (CAA-SSLA).
Peripheral clusters that are connected to the central core by only a few relationships represent the sole (or perhaps initial) integration of lone developers into a broader community.

@@

The highest mean centrality is seen in repositories that contain lists/datasets, standalone software, or packages/libraries, and in repositories based on Python, R or QGIS [@fig-repo-betweenness-category-platform].
Interestingly, these trends depart from the observed popularity of different categories/platform in the open-archaeo dataset as a whole (see @sec-open-archaeology): standalone software is more central than packages/libraries, even though there are more of the latter by a significant margin.
This may be due to the fact that many packages are developed to support specific practices or use-cases (often inspired by personal need), or are designed to run relatively stable statistical functions that need not change over time.
These are therefore relatively stable and require little additional input after release.
On the other hand, as discussed in @sec-open-archaeology, standalone software tend to integrate multiple system components and may evolve over time to add new features or support new workflows.
Moreover, standalone software are generally rooted in longer-term and community-held objectives, and their development may therefore be backed by insitutions with funding and resources to support developers.

Despite being the minority language, Python packages tend to be more central than R packages, on average.
This might relate to the different ways in which R and Python developers call dependencies ???
<!-- @ZB: I think this is interesting, but I couldn't really think of a general observation to make here. Maybe you can? -->
<!-- I wonder what the median is? Also: cltk is a python script, and is the massive pink blob in the repo-repo and user-user networks, so that might skew things slightly -->
